---
title: "Storm Impact Analysis in the US"
output: 
  html_document:
    highlight: pygments
    keep_md: true
    toc: true
    theme: cerulean
---

# Synopsis

This reports looks at the impact of storm events across the United States 
between the year 1950 and Novermber 2011. In particular, it considers the cost 
of these storm events both to population health and the economy. The analysis 
will show that the top 3 events that have been the most harmful to population 
health are _TORNADO_, _EXCESSIVE HEAT_ and _FLASH FLOOD_ taking into
account the number of fatalities and injuries caused by each event type.
Also, the top 3 events that have had the greatest economic consequences are 
_FLOOD_, _HURRICANE/TYPHOON_ and _TORNADO_ considering the cost (in dollars) of
damages to properties and crops.


# Data Processing
So as to ascertain the most harmful events with respect to population health, 
this report considered the following variables

-  the type of event (indicated in the EVTYPE variable)
-  number of deaths caused by an event (indicated in FATALITIES variable)
-  number of injuries caused by an event (indicated in INJURIES variable)

So as to ascertain the events that resulted in the gravest economic impact, the
following report took into account

- the type of event indicated in the EVTYPE variable
- dollar estimate of damage to properties indicated in the "PROPDMG" variable
- magnitude of damage estimate to properties indicated in the "PROPDMGEXP" variable
- dollar estimate of damage to properties indicated in the "CROPDMG" variable
- magnitude of damage estimate to properties indicated in the "CROPDMGEXP" variable

### Tidy up data

The variables mentioned above are extracted and cleaned up to allow further 
analysis required to answer the questions this report aims to address.

```{r, cache = TRUE}
storm <- read.csv('~/work/R/coursera/repdata-012/PeerAssessment2/StormData.csv')

# Extract relevant variables needed for computation
extract <- storm[,c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")]

# Tidy up PROPDMGEXP and CROPDMGEXP variables to take care of the various 
# representation of magnitude of damage estimate.
# (""  "-" "?" "+" "0" "1" "2" "3" "4" "5" "6" "7" "8" "B" "h" "H" "K" "m" "M")
tidyExpData <- function(x) {
    x <- as.character(x)
    x[x == ''] <- 0
    x[(x == 'h') | (x == 'H')] <- 2
    x[(x == 'k') | (x == 'K')] <- 3
    x[(x == 'm') | (x == 'M')] <- 6
    x[(x == 'b') | (x == 'B')] <- 9
    x[(x == '-') | (x == '+') | (x == '?')] <- 1
    x 
}

extract$PROPDMGExpValue <- tidyExpData(extract$PROPDMGEXP)
extract$CROPDMGExpValue <- tidyExpData(extract$CROPDMGEXP)
```



-   Assuming most harmful events with respect to population health are events 
    that caused fatalities or injuries, let us compute the total number of 
    fatalities and/or injuries per event type.
    
```{r, cache = TRUE}
library(plyr)
# Compute total number of fatalities and injuries per event type
evHarmAgg <- ddply(extract, "EVTYPE", summarise, 
                   total.fatalities = sum(FATALITIES), 
                   total.injuries = sum(INJURIES))

# Get a subset of events that caused fatalities or injuries
evHarm <- evHarmAgg[evHarmAgg$total.fatalities > 0 | evHarmAgg$total.injuries > 0, ]
evHarm <- evHarm[order(evHarm$total.fatalities, evHarm$total.injuries, decreasing=T), ]
```


-   Taking into account the PROPDMG, PROPDMGEXP, CROPDMG and CROPDMGEXP variables,
    let us compute the cost (in dollars) of damages to properties and crops per 
    event type

```{r, cache = TRUE}
library(plyr)
# Compute cost of damages to properties
extract$PROPDMGCOST <- extract$PROPDMG * 10^as.numeric(extract$PROPDMGExpValue)

# Compute cost of damages to crops
extract$CROPDMGCOST <- extract$CROPDMG * 10^as.numeric(extract$CROPDMGExpValue)

# Compute total cost of damages to properties and crops per event type
evCostAgg <- ddply(extract, "EVTYPE", summarise,
                   total.propdmgcost = sum(PROPDMGCOST),
                   total.cropdmgcost = sum(CROPDMGCOST))

# Get a subset of events that caused damages to properties and/or crops
evCost <- evCostAgg[evCostAgg$total.propdmgcost > 0 | evCostAgg$total.cropdmgcost > 0, ]
evCost <- evCost[order(evCost$total.propdmgcost, evCost$total.cropdmgcost, decreasing=T), ]
```

# Results
## Most harmful events with respect to population health

This section answers the question:

> Across the United States, which types of events (as indicated in the EVTYPE 
  variable) are most harmful with respect to population health?
    
```{r}
# Top 20 most harmful events with respect to population health
evHarm20 <- head(evHarm, 20)
evHarm20 

# Number of most harmful events
length(evHarm$EVTYPE)

# Total number of event types
length(evHarmAgg$EVTYPE)
```

From the results of analysis carried out, it can be observed that:

-   out of the `r length(evHarmAgg$EVTYPE)` event types in the data set
    `r length(evHarm$EVTYPE)` were the most harmful to  human population health
    in the US.
-   The top 5 most harmful events includes 
    + `r head(evHarm$EVTYPE, 5)`
    
A complete list of the most harmful events with respect to population health are
can be seen in the Appendix section below.

**Plot of the 20 most harmful events**

```{r}
library(reshape2)
evHarm20Melt <- melt(evHarm20, id=c('EVTYPE'), 
                     measure.vars=c('total.fatalities', 'total.injuries'))

library(lattice)
barchart(EVTYPE ~ value | variable, data = evHarm20Melt)
```


## Economic consequences

This section addresses the question:

> Across the United States, which types of events have the greatest economic 
  consequences?


```{r}
# Top 20 most harmful events
evCost20 <- head(evCost, 20)
evCost20
```

From the results computed above, which took into account the cost of damage to
both properties and crops, it can be seen that the top 5 events that have the
greatest economic consequences includes 

- `r head(evCost20$EVTYPE, 5)`.

**Plot of the events that caused the greatest economic consequence**

```{r}
library(reshape2)
evCost20Melt <- melt(evCost20, id=c('EVTYPE'), 
                     measure.vars=c('total.propdmgcost', 'total.cropdmgcost'))
barchart(EVTYPE ~ value | variable, data = evCost20Melt)
```


# Appendix

```{r}
# List of most harmful events
as.character(evHarm$EVTYPE)
```